Single-cell virus sequencing of influenza infections that trigger innate immunity

### 2018-10-06 paper/paper.tex
Minor tweaks that slightly shorten 3rd section of Results

We obtained transcriptomes for 1,614 human (A549) cells, and 50 of the uninfected canine cells that were spiked into the experiment as a control (\FIG{transcriptomics}A).
We also obtained 12 transcriptomes with a mix of human and canine transcripts; these mixes arise when multiple cells are captured in a single droplet.
From the number of mixed cell-type transcriptomes, we estimate~\citep{bloom2018estimating} that $\sim$11\% of the transcriptomes are derived from multiple cells.
To remove some of these multiplets, we filtered transcriptomes with unusually high or low numbers of cellular transcripts (\FIG{transcriptomics}B).
For the remainder of this paper, we analyze the 1,490 human cells that passed this filtering.

To identify infected cells, we examined the fraction of each transcriptome derived from virus (\FIG{transcriptomics}B).
As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were from influenza; this low-level background is likely from lysed cells that release ambient viral mRNA that is captured in droplets during reverse-transcription.
We tested whether each cell contained significantly more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 of the 1,490 human cells as definitively infected with influenza (\FIG{transcriptomics}B).
We classified the other cells as uninfected, although it is possible that some were infected with virions that produced too little mRNA to be detected in our experiment.
The distribution of the amount of viral mRNA across infected cells is shown in the inset in \FIG{transcriptomics}B.
As in our prior work~\citep{russell2018extreme}, the distribution is extremely heterogeneous: many infected cells have only a few percent of mRNA derived from virus, but viral mRNA comprises over half the transcriptome of a few cells.

We then called the presence or absence of each viral gene in each infected cell, again using a Poisson model parameterized by background fractions estimated from uninfected canine cells.
Although ten transcripts are encoded by the eight viral genes, we call presence / absence of genes rather than transcripts.
The reason is that the genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable by 3'-end sequencing.
\FIG{transcriptomics}D (top panel) shows that most (162 of 290) infected cells express all eight genes, although a substantial minority fail to express one or more genes (see \FIGSUPP[transcriptomics]{frac_has_gene} for frequencies at which individual genes are present).
This measured frequency of infected cells expressing all eight genes is higher than in our own prior work~\citep{russell2018extreme} and studies by others~\citep{brooke2013most, heldt2015single, dou2017analysis}, which have estimated that only 13\% to 50\% of infected cells express all genes. 
However, it is important to remember that the cells in our experiment have been enriched for IFN expression by MACS sorting, which could alter the distribution of viral gene expression from a fully random sample of infected cells.

The amount of viral mRNA tended to be lower in cells that failed to express viral genes (\FIG{transcriptomics}D, bottom panel), consistent with our prior work~\citep{russell2018extreme}.
However, viral burden remained highly variable even after conditioning on the number of viral genes: some cells that failed to express one or even two genes still derived over 50\% of their mRNA from influenza, while other cells that expressed all viral genes had only a few percent of their mRNA from influenza (\FIG{transcriptomics}D, bottom panel).
Also consistent with our prior work~\citep{russell2018extreme}, despite the wide variation in absolute expression of viral genes, their \emph{relative} expression was fairly consistent across cells (\FIG{transcriptomics}E) and generally matched values obtained by older bulk studies~\citep{hatada1989control}.

By examining the synonymous viral barcodes near the 3' termini of transcripts, we determined that 38\% of cells were co-infected with wild-type and synonymously barcoded virions (\FIG{transcriptomics}F; cells called as co-infected if a binomial test rejected null hypothesis that $\ge$95\% of viral mRNA is from one viral barcode variant).
From \FIG{transcriptomics}F, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode (such co-infections cannot be identified from the transcriptomic data).
Interestingly, this co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is Poisson.
This discrepancy could arise if the MACS for IFN+ cells also enriches co-infected cells, if infection itself is not truly Poisson, or if co-infection increases the likelihood that we identify a cell as infected given the thresholds in \FIG{transcriptomics}C.
This moderately high rate of co-infection may also partially explain why more cells in our experiment express all eight viral genes compared to some prior studies, as a co-infecting virion can complement a missing viral gene~\citep{russell2018extreme}.

>>>>

We obtained transcriptomes for 1,614 human (A549) cells, and 50 of the uninfected canine cells that were spiked into the experiment as a control (\FIG{transcriptomics}A).
We also obtained 12 transcriptomes with a mix of human and canine transcripts; from the number of such mixed cell-type transcriptomes, we estimate~\citep{bloom2018estimating} that $\sim$11\% of the transcriptomes are derived from multiple cells.
To remove some of these multiplets, we filtered transcriptomes with unusually high or low numbers of cellular transcripts (\FIG{transcriptomics}B), retaining the 1,490 human cells that passed this filtering.

To identify infected cells, we examined the fraction of each transcriptome derived from virus (\FIG{transcriptomics}B).
As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were from influenza; this low-level background is likely from lysed cells that release ambient viral mRNA that is captured in droplets during reverse-transcription.
We tested whether each cell contained significantly more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected with influenza (\FIG{transcriptomics}B).
We classified the other cells as uninfected, although it is possible that some were infected with virions that produced too little mRNA to be detected in our experiment.
The distribution of the amount of viral mRNA across infected cells is shown in the inset in \FIG{transcriptomics}B.
As in our prior work~\citep{russell2018extreme}, the distribution is extremely heterogeneous: many infected cells have only a few percent of mRNA derived from virus, but viral mRNA comprises over half the transcriptome of a few cells.

We then called the presence or absence of each viral gene in each infected cell, again using a Poisson model parameterized by background fractions estimated from uninfected canine cells.
We called presence / absence of genes rather than transcripts, since the two influenza genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable by 3'-end sequencing.
\FIG{transcriptomics}D (top panel) shows that most (162 of 290) infected cells express all eight genes, although a substantial minority fail to express one or more genes (see \FIGSUPP[transcriptomics]{frac_has_gene} for frequencies at which individual genes are present).
This measured frequency of infected cells expressing all eight genes is slightly higher than in our own prior work~\citep{russell2018extreme} and studies by others~\citep{brooke2013most, heldt2015single, dou2017analysis}, which have estimated that only 13\% to 50\% of infected cells express all genes.

The amount of viral mRNA tended to be lower in cells that failed to express viral genes (\FIG{transcriptomics}D, bottom panel).
However, viral burden remained highly variable even after conditioning on the number of viral genes: some cells that failed to express one or even two genes still derived $>$50\% of their mRNA from virus, while other cells that expressed all genes had only a few percent of their mRNA from virus (\FIG{transcriptomics}D, bottom panel).
Consistent with our prior work~\citep{russell2018extreme}, despite the wide variation in absolute expression of viral genes, their \emph{relative} expression was fairly consistent (\FIG{transcriptomics}E) and generally matched values from older bulk studies~\citep{hatada1989control}.

By examining the synonymous viral barcodes near the 3' termini of transcripts, we determined that 38\% of cells were co-infected with wild-type and synonymously barcoded virions (\FIG{transcriptomics}F; cells called as co-infected if a binomial test rejected null hypothesis that $\ge$95\% of viral mRNA is from one viral barcode variant).
From \FIG{transcriptomics}F, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode (such co-infections cannot be identified from transcriptomic data).
Interestingly, this co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is Poisson.
This discrepancy could arise if the MACS for IFN+ cells also enriches co-infected cells, if infection is not truly Poisson, or if co-infection increases the likelihood that we identify a cell as infected given the thresholds in \FIG{transcriptomics}C.
This moderately high rate of co-infection may also explain why more cells in our experiment express all eight viral genes compared to some prior studies, as a co-infecting virion can complement a missing viral gene~\citep{russell2018extreme}.
